RCloud: Integrating Exploratory Visualization, Analysis and Deployment

### 2014-03-27 discussion.tex
yes

\section{Lessons Learned, Discussion and Limitations}

@@

Experience with deploying RCloud in a community of data analysts
has provided some insight into how well the proposed requirements
satisfied their needs. This may be informative to researchers and
designers of future systems supporting visual analytics.

@@

{\bf Shared artifacts and artifact histories} are a central feature.
The prototype aimed at making it convenient to collaborate on code
and experiments with data. It was a success to observe that analysts
readily shared ideas through RCloud. We observed them creating experiments
in workbooks and sharing URLs to discuss analysis and programming
problems, instead of sending screenshots in Powerpoint.

{\bf View sharing, bookmarking.} Likewise, everything is named
and accessed as a URL. This turned out to be an effective mechanism
to support sharing and integration of analyses in notebooks
with external processes and systems.

{\bf Discussion.} Annotation is supported in our human interface,
and was also a central goal for our prototype. \todo{Elaborate.}
An interesting question is to what extent users of production notebooks
should be able to use tools like annotation without coding and
without being exposed to the analyst's view.\todo{Is this a gap in
our design?}

{\bf Content export} is not a capability we worked to support,
but the R environment already has many packages for this.
Our goal was to devise ways of working where explicit export is
less necessary. However it is desirable to integrate results
through web protocols such as JSON. We did assume that RCloud
notebooks need to play well in the ecosystem of other tools,
and this worked out as intended, though we learned it also creates
a temptation to absorb features almost without limit. (Should
RCloud support templates for personal web pages? Should it
incorporate emacs on the web, and an email client?)

{\bf Social-psychological incentives} and {\bf voting and ranking}
are supported through starring. This mechanism was employed often
by our test audience.  An obvious next step would be to enhance
recommendations using relationships discovered by static and
dynamic code analysis. This should be done both within and across
collections of scripts (the latter being similar to work in Vistrails
that enhanced recommendations by clustering multiple workflows).
It may be valuable to data analysts to know which packages are often
used together, or with a given type of data. It is probably helpful
to know which data sources and packages often go together. In general,
though, trivial or even passive mechanisms to collect data for
recommendations are essential.

{\bf Group management, size and diversity.} This area lacks sufficient
support in RCloud, but is clearly desirable. We rely on conventional
external administrative and social processes, outside our platform,
to manage accounts and groups.  Automation of this capability in a
large organization could rely on social media services

@@

Because we developed the ideas and the prototype together,
we could not forsee some of the requirements that emerged
after our audience started working with the prototype.

It was unexpected that we had to had to back off so soon on
the assumption that programming is being done in only one
language (R). In retrospect, it should have been obvious that
many data science problems are already being solved with
cross-language approaches.

A clear disdavantage is that some important aspects of the environment in
RCloud are more difficult to manage than in conventional systems. Not only
do we lack an explicit separation of the development and deployment
environments, but every version of every prototype is shared by default.
While this encourages collaboration on work in progress, errors such as
misconfiguration, mismatches between versions of packages,or ordinary
programming errors are also easily shared and could affect production
websites unintentionally. We take the position that this cannot entirely
be avoided, but the power of sharing is worth this price (just as that
has proven true with distributed systems in general). Still, this 
problem needs more attention. 
Versioning occurs in several places: in scripts, in the R environment
such as the installed R libraries and packages, and in the external
environent such as the operating system and even protocols spoken by
remote services. At one end of the scale, we have full control over the
versioning of scripts via git, though there is still a need for conventions
for example to name stable or working versions of scripts or versions known
to work together. The R environment is under the control of its package
manager. It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a conservative approach to upgrades.

On top of this, RCloud allows analysts to fork experiments to try
new ideas quickly. But we have not done much so far to address the
probem of having a lot of bits and pieces of code and data lying
around, though at least we have made it easier to get at them and
to apply algorithms and metadata to organize them.

@@

Our environment assumes a high level of trust within the user community.
Access control for information security is implemented by relying on the
host operating system and web server. We admit this may require further

>>>>

\section{Discussion, Lessons and Limitations}

@@

Experience with deploying RCloud in a community of data analysts
has provided some insight into how well the proposed requirements
were met and matched their work. This may be informative to
researchers and designers of future systems supporting visual analytics.

@@

{\bf Shared artifacts and artifact histories} are a central feature
of RCloud. The prototype aimed at making it convenient to collaborate
on code and experiments with data. It was a success to observe that analysts
readily shared ideas through RCloud. We noticed them creating experiments
in workbooks and sharing URLs to discuss analysis and programming
problems, instead of sending screenshots in PowerPoint.\todo{Talk about artifact histories?}

{\bf View sharing, bookmarking.} Almost everything in RCloud is named
and accessed as a URL. This proved to be an effective mechanism
for sharing, and for analyses in notebooks to be integrated with
external processes and systems.

{\bf Discussion.} Annotation is supported in our human interface,
and was another central goal of our prototype. \todo{Elaborate.}
An interesting question is, to what extent should users of production
notebooks also able to use annotation and similar tools without coding
and without being exposed to the analyst's view?\todo{Is this a gap in
our design?}

{\bf Content export} is not a capability we worked to support,
but the R environment already has many packages for this.
Our goal was to devise ways of working where explicit export
is less necessary. However it is desirable to integrate results
through web protocols such as JSON. We did assume that RCloud
notebooks need to play well in the ecosystem of other tools,
and this worked out as intended, though we learned it also creates
a temptation to absorb features almost without limit. (Should
RCloud support templates for personal web pages and graphical
editing? Should it incorporate an email client and emacs?
Should it be able to push results to Facebook?)

{\bf Social-psychological incentives} and {\bf voting and ranking}
are supported through starring. This mechanism was employed often
by our test audience. An obvious next step would be to enhance
recommendations via relationships discovered by static and
dynamic code analysis. This should be done both within and across
collections of scripts (the latter being similar to Vistrails'
enhanced recommendations by clustering multiple workflows).
It should be valuable to data analysts to know which packages are
frequently used together, or appear with a given record type
or data feed. In general, though, trivial or passive mechanisms
to collect data for recommendations are essential.

{\bf Group management, size and diversity.} This area lacks sufficient
support in RCloud, but is clearly important to working in teams.
We rely on existing administrative and social processes and conventions,
outside RCloud, to manage accounts and groups. This capability
could likely benefit from integrating social media to track
identities and groups and to maintain communication channels.

@@

Because we developed the ideas and the prototype together,
we did not foresee some of the requirements that emerged
after our audience started working with the prototype.

We did not anticipate having to back off so soon on the
assumption that programming is being done mainly in one
language (R). In retrospect, it should have been obvious
that many data science problems are already being solved
with cross-language approaches.

A disadvantage is that some important aspects of the environment
are more difficult to manage in RCloud than in conventional systems.
Not only does it lack an explicit separation of the development and
deployment environments, but every version of every prototype is
shared by default. While this encourages collaboration on work
in progress, errors such as misconfiguration, mismatches between
versions of packages, and ordinary programming errors are also
easily shared and can affect production websites unintentionally.
This cannot be avoided entirely, but the power of sharing is worth
this price (as proven true with distributed systems in general).
Still, this problem needs more attention if the environment is
going to be scaled up to a wider number of users with fewer
informal channels to coordinate work. \todo{need a transition}
Versioning occurs in several places: in scripts, in the R environment
such as the installed R libraries and packages, and in the external
environment such as the operating system and even protocols spoken by
remote services. At one end of the scale, we have full control over the
versioning of scripts via git, though there is still a need for conventions
for example to name stable or working versions of scripts or versions known
to work together. The R environment is under the control of its package
manager. It is at least possible for RCloud to access information about
package versions and configurations, but support for checking compatibility
and maintaining multiple versions in the same environment is not strong.
At the other end of the scale, there is not much reason to expect version
control for the external environment. Most programmers rely on clean living
and a conservative approach to upgrades.

On top of this, RCloud allows analysts to fork experiments to try
new ideas quickly. But we have not done much so far to address the
problem of having a lot of bits and pieces of code and data lying
around, though at least we have made it easier to get at them and
to apply algorithms and metadata to organize them.

@@

Our environment assumes a high level of trust between its users.
Access control for information security is deferred to the host
operating system and web server. This decision should be re-examined,
but more sophisticated approaches such as information flow analysis
are an active research area and many subtle problems remain
\cite{Moore:2011:SAF}. We are hopeful that providing a richer
environment for collaboration and information sharing may aid
the application of information security research.

### 2014-03-27 relatedwork.tex
yes

VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
are tools built on top of the VisTrails workflow management
system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website feature
workflow upload and remote execution. In our view, the impedance
mismatch between a dataflow pipeline specification and the power of a
general-purpose language is too great for the type of general
exploratory work in data science teams. As a result, RCloud tries to
provide a closer data-analysis to analysts accustomed to createing and
executing R and Python code while keeping many of the attractive
properties like automatic versioning and management.

Our work has been inspired by and benefited from other proposals
to improve data analysis environments and processes,
including RStudio \cite{RStudio:2013:SWA},
R packages such as Markdown \cite{Allaire:2014:MMR},
knitr \cite{Xie:2013:DDW}
and Shiny \cite{RStudio:2013:SWA},
and iPython notebooks \cite{Perez:2007:IAS}
to name a few. RStudio aims at providing an integrated development environment
for R programming, with support for publishing code in packages. Markdown,
knitR and Shiny provide R with sophisticated reporting capabilities, including
interactive web interfaces. IPython \cite{Perez:2007:IAS}
shares many of our goals, such as providing a comprehensive environment
for analysis and programming, with sharable documents on the web.

One overall goal is to reduce the gap between implementors and deployers
of technology. Don't want a team of 20 IT people to support 5 data scientists.

>>>>

VisMashup~\cite{Santos:2009:VST} and Crowdlabs~\cite{Mates:2011:CSA}
are tools built on top of the VisTrails workflow management
system~\cite{Callahan:2006:VVM}. VisMashup defines a schema and
semantics for automatically deriving user interfaces from workflows,
while Crowdlabs exposes these capabilities on a website feature
workflow upload and remote execution. In our view, the impedance
mismatch between a dataflow pipeline specification and the power of a
general-purpose language is too great for the type of general
exploratory work in data science teams. As a result, RCloud tries to
provide a closer data-analysis to analysts accustomed to creating and
executing R and Python code while keeping many of the attractive
properties like automatic versioning and management.

Our work has been inspired by and benefited from other proposals
to improve data analysis environments and processes,
including RStudio \cite{RStudio:2013:SWA},
R packages such as Markdown \cite{Allaire:2014:MMR},
knitr \cite{Xie:2013:DDW}
and Shiny \cite{RStudio:2013:SWA},
and iPython notebooks \cite{Perez:2007:IAS}
to name a few. RStudio aims at providing an integrated development environment
for R programming, with support for publishing code in packages. Markdown,
knitR and Shiny provide R with sophisticated reporting capabilities, including
interactive web interfaces. IPython \cite{Perez:2007:IAS}
shares many of our goals, such as providing a comprehensive environment
for analysis and programming, with shareable documents on the web.

One overall goal is to reduce the gap between implementers and deployers
of technology. Don't want a team of 20 IT people to support 5 data scientists.

### 2014-03-27 system.tex
yes

The internal computing infrastructure of organizations has changed
radically in the last fifteen years. The shift from large
servers toward scalable, lower-cost, distributed systems (``the cloud'')
led to a software ecosystem of processes distributed over a
network (often virtually defined) communicating via some
high-level protocol, usually HTTP.  HTTP is dominant because web
browsers and servers are ubiquituous and available in essentially
all hardware devices, from tiny sensors, to handheld devices and
laptops, to rack-mounted servers.

@@

We incorporate both explicit and implicit indications of interest
in notebooks. Explicit interest is signalled by ``starring,'' or
clicking on a button that marks a notebook as interesting. 
This makes explicit indication of interest a nearly trivial operation,
always readily available, thus encouraging its use.

@@

There exists a URL for every notebook in RCloud, and notebooks by
default are visible by the entire organization. This is deliberate.
As pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS},
broad access to analysis outputs (in their case, for NameVoyager)
increases long-term engagement in part through crossreferences on
the web. Although our prototype RCloud deployment is only visible
inside a corporate intranet, we nevertheless found anecdotal support
for this notion by discovering links to RCloud notebooks in internal
discussion fora and mailing lists.

>>>>

The internal computing infrastructure of organizations has changed
radically in the last fifteen years. The shift from large
servers toward scalable, lower-cost, distributed systems (``the cloud'')
led to a software ecosystem of processes distributed over a
network (often virtually defined) communicating via some
high-level protocol, usually HTTP.  HTTP is dominant because web
browsers and servers are ubiquitous and available in essentially
all hardware devices, from tiny sensors, to handheld devices and
laptops, to rack-mounted servers.

@@

We incorporate both explicit and implicit indications of interest
in notebooks. Explicit interest is signaled by ``starring,'' or
clicking on a button that marks a notebook as interesting. 
This makes explicit indication of interest a nearly trivial operation,
always readily available, thus encouraging its use.

@@

There exists a URL for every notebook in RCloud, and notebooks by
default are visible by the entire organization. This is deliberate.
As pointed out by Wattenberg and Kriss~\cite{Wattenberg:2011:DFS},
broad access to analysis outputs (in their case, for NameVoyager)
increases long-term engagement in part through cross-references on
the web. Although our prototype RCloud deployment is only visible
inside a corporate intranet, we nevertheless found anecdotal support
for this notion by discovering links to RCloud notebooks in internal
discussion fora and mailing lists.
